---
title: "Chlamydia in Benton County"
author: "Keerthana Radhakrishnan"
date: "4/18/2020"
output: word_document
---
Chlamydia is a common sexually transmitted infection which is caused by bacteria named Chlamydia trachomatis.
Common symptoms in women include: abnormal vaginal discharge, a burning sensation when urinating, and pain during intercourse.
In men, discharge from the penis, burning sensation when urinating, and burning or itching around the opening of penis are some of the symptoms.

Reading the data into R Studio:
```{r}
setwd("~/Downloads/internship/Project BCHD/Chlamydia") # setting word directory
ct<-read.csv("Chlamydia data2.csv", as.is=T, sep=",", header=T, na.strings = TRUE)
View(ct) # to view data in a separate window
```
This data analysis requires the following R packages:
1. lubridate
2. dplyr
3. tidyr
4. ggplot2
5. lattice

```{r}
ct$CollectionDate<-mdy(ct$EXP_Lab..CollectionDate) #Converting "EXP_Lab..CollectionDate" variable into a date variable named "CollectionDate" using Lubridate package

#Creating separate variables for year, month and date 
ct = ct %>% 
  mutate(CollectionDate = ymd(CollectionDate)) %>% 
  mutate_at(vars(CollectionDate), funs(year, month, day))%>%
  filter(year>=2010 & year<2020) # filtering data between 2010 and 2019


#Creating new variable "count" which counts the number of cases per month for each year.
cases<- ct%>%
  group_by(year,month)%>%
  summarise(count=n())
```

Trellis plot: the following trellis plot represents the number of cases against month per year (2010-2019):
Using Lattice package to create this plot
```{r}
xyplot(cases$count~cases$month|factor(cases$year),layout=c(3,4),main="Trellis plot: cases per month from 2010-2019",type='b',xlab = "month", ylab = "count",aspect = "fill",strip = TRUE,col="blue")
```
```{r}
# Creating wide format of the "cases" dataset:
year_wide<-spread(cases,month,count)

# creating a new variable "tot_cases" for summing up the number of cases per year
for (i in year_wide$year){
  year_wide$tot_cases <- rowSums(year_wide[2:13],na.rm=TRUE)
  print(year_wide)
}

# including a new variable "tot.pop" for total population per year in Benton county
year_wide$tot.pop<- c(85735,85995,86785,87725,88740,90005,91320,92575,93590,94360)

# creating new variable "rate" and including the rates per 100,000 population in this variable for each year
year_wide<- year_wide%>%
  mutate(rate=((tot_cases/tot.pop)*100000))

# Including two new variables the records no. of cases in Oregon and their rates
year_wide$OR_cases<-c(12342,13687,	13493,	14255,	15338,	16336,	17417,	18641,	19103,	19236)
year_wide$OR_rates<- c(322,	355,	347,	364,	387,	407,	427,	450,	455,	454)

# to take a look at the data
year_wide
```

5-year average of Chlamydia (2015-2019):
```{r}
five_avg<- year_wide %>%
  filter(year<=2019 & year>=2015) # creating new dataset for years 2015-2019); I can also use "filter" and "mutate" options for this but I prefer using basic R

# For Benton County,
mean(five_avg$tot_cases)
mean(five_avg$rate)

# For Oregon
mean(five_avg$OR_cases)
mean(five_avg$OR_rates)
```
10- year average (2010-2019):
```{r}
# Benton county
mean(year_wide$tot_cases)
mean(year_wide$rate)

# For Oregon
mean(year_wide$OR_cases)
mean(year_wide$OR_rates)
```
Graphs:

1) Plot for number of cases per year; 
```{r}
# It is important to convert "year" to factor before plotting the data:
year_wide$year<-as.factor(year_wide$year)

# Creating plot for Benton county:
ggplot(year_wide,aes(x=year,y=tot_cases,group=1))+
  geom_line(colour="blue",lty=7,lwd=1)+
  geom_point(colour="blue",lwd=3)+
  theme_bw()+
  labs(x="Year", y="Number of cases", title = "Chlamydia cases in Benton county- 2010 to 2019")

# the same for Oregon:
ggplot(year_wide,aes(x=year,y=OR_cases,group=1))+
  geom_line(colour="darkgreen",lty=7,lwd=1)+
  geom_point(colour="darkgreen",lwd=3)+
  theme_bw()+
  labs(x="Year", y="Number of cases", title = "Chlamydia cases in Oregon- 2010 to 2019")

```
Comparing Chlamydia rates in Oregon and Benton county between 2010 and 2019:
```{r}
year_wide$year<-as.factor(year_wide$year)
colors<-c("Benton county"="blue","Oregon"="darkgreen")
ggplot(year_wide,aes(x=year,y=rate,group=1))+
  geom_line(aes(x=year,y=rate,group=1,colour="Benton"),lwd=1)+
  geom_line(aes(x=year,y=OR_rates,group=1,colour="Oregon"),lwd=1)+
  geom_point(aes(x=year,y=rate,group=1),colour="blue",lwd=2)+
  geom_point(aes(x=year,y=OR_rates,group=1), colour="darkgreen",lwd=2)+
  scale_y_continuous(breaks=seq(0,700,100))+
  theme_bw()+
  labs(x="Years", y="rate (per 100,000)", title = "Chlamydia rates in Benton County and Oregon from years 2010-2019",color="Legend")+
  scale_colour_manual(name="Legend",values=c(Benton="blue", Oregon="darkgreen"))
```
Interpretation: As we can see above, there is a huge difference in the rates of chlamydia between Oregon and Benton county, with county having a lot more cases than the state.

Chlamydia and gender:

```{r}
#Creating a new dataset "gender" which includes year, Sex and number of cases per gender for each year:
gender<- ct%>%
  group_by(year,Sex)%>%
  summarise(number=n())
# gender<- gender[which(gender$year<2020 & gender$year>=2010),]
# removing missing values:
gender[which(gender$Sex==""),]<-NA
gender<-na.omit(gender)

#take a look at the new dataset:
gender
```
Graphs:
1) Pyramid plot for number of cases against year categorized by gender:
```{r}
ggplot(data=gender) +
  geom_bar(aes(year,number,group=Sex,fill=Sex), stat = "identity",subset(gender,Sex=="F")) +
  geom_bar(aes(year,-number,group=Sex,fill=Sex), stat = "identity",subset(gender,Sex=="M")) +
  scale_y_continuous(breaks=seq(-1000,1000,50),labels=abs(seq(-1000,1000,50))) +
  scale_x_continuous(breaks= seq(2010,2019,1))+
  theme_bw()+
  labs(x="Year",y="Number of cases",title="Chlamydia cases per year for males and females (2010-2019)")+
  coord_flip() +
  scale_fill_discrete(name="Gender", breaks=c("F", "M"),labels=c("Female","Male"))
```
2) Scatterplot for number of cases based on gender:
```{r}
gender$year<-as.factor(gender$year)
ggplot(gender,aes(x=year,y=number,group=Sex))+
  geom_line(aes(col=Sex),lwd=2)+
  labs(x="Year",y="Number of cases",title="Chlamydia cases in Benton county by gender- 2010 to 2019")+
  scale_colour_discrete(name="Gender", breaks=c("F", "M","U"),labels=c("Female","Male","Unidentified"))+
  theme_bw()
```
Chlamydia and age:
```{r}
# creating new dataset that contains year, Age, Sex, and the count of the cases:
ct_age_cat<- ct%>%
  group_by(Age,Sex)%>%
  summarise(counts=n())

# Creating a new variable " age_cat" categorising age into age groups:
ct_age_cat<- ct_age_cat%>%
  mutate(age_cat=case_when(Age < 10 ~ "(0-9)",
                           Age >= 10  & Age <= 19 ~ "(10-19)",
                           Age >= 20  & Age <= 29 ~ "(20-29)",
                           Age >= 30  & Age <= 39 ~ "(30-39)",
                           Age >= 40  & Age <= 49 ~ "(40-49)",
                           Age >= 50  & Age <= 59 ~ "(50-59)",
                           Age >= 60  & Age <= 69 ~ "(60-69)",
                           Age >= 70  & Age <= 79 ~ "(70-79)",
                           Age >= 80  & Age <= 89 ~ "(80-89)",
                           Age >= 90  & Age <= 99 ~ "(90-99)"))

ct_age_cat # take a look at the new dataset
```

Graph showing chlamydia cases by age group and gender:
```{r}
#ct_age_cat$counts<-as.integer(ct_age_cat$counts,na.rm=TRUE)
ggplot(ct_age_cat,aes(counts,group=Sex,fill=Sex))+
  geom_bar(aes(age_cat,counts,group=Sex,fill=Sex), stat = "identity",subset(ct_age_cat,Sex=="F"),na.rm=TRUE) +
  geom_bar(aes(age_cat,-counts,group=Sex,fill=Sex), stat = "identity",subset(ct_age_cat,Sex=="M"),na.rm=TRUE) +
  scale_y_continuous(breaks=seq(-1000,4000,200),labels=abs(seq(-1000,4000,200))) +
  scale_x_discrete()+
  theme_bw()+
  labs(x="Age range",y="Number of cases",title="Chlamydia in Benton county by Age groups and Gender")+
  coord_flip()+
  scale_fill_discrete(name="Gender", breaks=c("F", "M"),labels=c("Female","Male"))
```
**Interpretation:**
The above graph shows the number of cases recorded for different age groups categorised by gender. 
The graph shows that gonorrhoea cases are higher in females than in males and highest number of cases are found in persons between 20-29 years age irrespective of the gender.

Chlamydia and race:
```{r}
# creating new dataset for year, race and count of cases per race and year:
ct_race_cases<- ct%>%
  group_by(year,EXP_Patient..Race)%>%
  summarise(count=n())

table(ct_race_cases$EXP_Patient..Race) # gives a table of race in the dataset
```
Categorization of race:
"Two races"= Black\nAI/AN","Asian\n","Asian\nOther","Asian\nPacific Is.","Black\nOther", "Pacific Is.\nAsian","White\nAsian","White\nOther","White\nAI/AN","Other\n","Asian\nWhite",White\n

```{r}
# combining a few category of races to create a new "two races" category:
ct_race_cases$EXP_Patient..Race[which(ct_race_cases$EXP_Patient..Race=="Black\nAI/AN"|ct_race_cases$EXP_Patient..Race=="Asian\n" |ct_race_cases$EXP_Patient..Race=="Asian\nOther"|ct_race_cases$EXP_Patient..Race=="Asian\nWhite"|ct_race_cases$EXP_Patient..Race=="Asian\nPacific Is."|ct_race_cases$EXP_Patient..Race=="Black\nOther" |ct_race_cases$EXP_Patient..Race=="Pacific Is.\nAsian"| ct_race_cases$EXP_Patient..Race=="White\nAsian"|ct_race_cases$EXP_Patient..Race=="White\nOther"|ct_race_cases$EXP_Patient..Race=="White\nAI/AN"|ct_race_cases$EXP_Patient..Race=="Other\n"|ct_race_cases$EXP_Patient..Race=="White\n")] <- "Two races"

# Adding total population as a new variable: 
ct_race_cases<- ct_race_cases%>%
  mutate(total_pop = case_when(year == "2010" ~ 85735 ,
                           year == "2011" ~ 85995,
                          year == "2012" ~86785,
                          year == "2013" ~87725,
                          year == "2014" ~88740,
                          year == "2015" ~90005,
                          year == "2016" ~91320,
                          year == "2017" ~92575,
                          year == "2018" ~93590,
                          year == "2019" ~94360))

# to calculate rates per 100,000:
ct_race_cases$Rate<-c(((ct_race_cases$count)/(ct_race_cases$total_pop))*100000)

# assigning missing values and removing them from the dataset:
ct_race_cases$EXP_Patient..Race<-as.character(ct_race_cases$EXP_Patient..Race)
ct_race_cases[which(ct_race_cases$EXP_Patient..Race==""),]<- NA
ct_race_cases<-na.omit(ct_race_cases)

# to look at the edited dataset:
ct_race_cases
```
Graphs:
```{r}

#for number of cases per race:
ct_race_cases$year<-as.factor(ct_race_cases$year)
ggplot(ct_race_cases,aes(year,count,group=EXP_Patient..Race))+
  geom_line(aes(colour=EXP_Patient..Race),linetype=3)+
  geom_point(aes(colour=EXP_Patient..Race),size=2)+
  scale_color_discrete(name="Race")+
  labs(x="Year",y="Number of cases",title="Chlamydia cases categorised by Race (2010-2019)")+
  theme_bw()

# for rates per 100,000 population per race:
ct_race_cases$year<-as.factor(ct_race_cases$year)
ggplot(ct_race_cases,aes(year,Rate,group=EXP_Patient..Race))+
  geom_line(aes(colour=EXP_Patient..Race),linetype=3)+
  geom_point(aes(colour=EXP_Patient..Race),size=2)+
  scale_color_discrete(name="Race")+
  labs(x="Year",y="Rate (per 100,000)",title="Chlamydia rates per 100,000 categorised by Race")+
  theme_bw()
```
Lets remove "Refused" and "Unknown" to get the graph solely for races:
```{r}
Nct_race_cat$year<-as.factor(Nct_race_cat$year)
ct_race_cat[which(ct_race_cat$Race=="Ref"|ct_race_cat$Race=="" ),]<- NA
Nct_race_cat<-na.omit(ct_race_cat)
Nct_race_cat
ggplot(Nct_race_cat,aes(year,Rate,group=Race))+
  geom_line(aes(colour=Race),linetype=3)+
  geom_point(aes(colour=Race),size=2)+
  scale_color_discrete(name="Race", breaks=c("AI/","Asi","Bla","Oth","Pac","Two","Whi"),labels=c("AI/AN", "Asian","Black","Others","Pacific Is","Two races","White"))+
  labs(x="Year",y="Rate (per 100,000)",title="Chlamydia rates (per 100,000) in Benton county by Race- 2010-2019")+
  theme_bw()
```
Chlamydia and Seasonal variation:

Creating a new dataset with a new categorical variable "season_cases" that categorizes the "month" variable into fall, spring, winter and summer. Another variable "total" sums up the count for each season per year.  
```{r}
ct_season_var<-cases%>%
  mutate(season_cases= ifelse((month==12|month==1|month==2),"Winter",ifelse((month==3|month==4|month==5),"Spring",ifelse((month==6|month==7|month==8),"Summer",ifelse((month==9|month==10|month==11),"Fall","NA")))))%>%
  group_by(year,season_cases)%>%
  summarise(total= sum(count))

ct_season_var # to take a look at the data
```
Dealing with missing values and ommitting them:
```{r}
ct_season_var[which(ct_season_var$season_cases=="NA"),]<-NA
ct_season_var<-na.omit(ct_season_var)
```
Barplot for seasonality. For this, I am converting the "year" variable into factor for representation purposes. If I don't take this step, the year will be considered as an integer.
```{r}
ct_season_var$year<-as.factor(ct_season_var$year)
ggplot(data=ct_season_var,mapping=aes(x=year,y=total,group=season_cases,fill=season_cases))+
  geom_bar(stat="sum",na.rm=T,position = "stack",colour="white",lwd=0.5)+
  geom_text(aes(label=total),position = position_stack(0.5),size=3.5,colour="white")+
  scale_y_continuous()+
  theme_bw()+
  labs(x="Year",y="Number of cases ",title="Chlamydia cases in Benton county categorised by season (2010-2019)")+
  scale_fill_discrete(name="Seasons")
```

































