---
title: "Syphilis in Benton County"
author: "Keerthana Radhakrishnan"
date: "4/21/2020"
output: word_document
---
Setting th working directory and reading datafile:
```{r}
setwd("~/Downloads/internship/Project BCHD/Syphilis")
st<-read.csv("Syphilis_data.csv", header = TRUE, as.is=TRUE, sep=",",na.strings = c(" ","NA",NA))
View(st) # to view the dataset
```
Packages used:
```{r}
require(lubridate)
require(dplyr)
require(lattice)
require(ggplot2)
```

```{r}
# Converting "EXP_Lab..CollectionDate" variable into date format "CollectionDate"
st$CollectionDate<-mdy(st$EXP_Lab..CollectionDate)

# Creating three separate variables for year, month and date
st = st %>% 
  mutate(CollectionDate = ymd(CollectionDate)) %>% 
  mutate_at(vars(CollectionDate), funs(year, month, day)) %>%
  filter(year>=2010 & year<= 2019)
```
Creating an new dataset including year, month and count of cases per month for each year:
```{r}
st_cases<- st%>%
  group_by(year,month)%>%
  summarise(count=n())
# here, I asked R to provide me a summary count of cases grouped by year and month

# converting the data to wide format.

st_cases_wide<-spread(st_cases,month,count)
View(st_cases_wide)  
```
From the wide format data, we can see that the cases of syphilis has increased in 2018 and 2019. 

Trellis plot:
```{r}
xyplot(st_cases$count~st_cases$month|factor(st_cases$year),layout=c(3,4),main="Trellis plot: cases per month from 2010-2019",type='b',xlab = "month", ylab = "count",aspect = "fill",strip = TRUE,col="blue")
```
The plot above adds evidence to the the finding that the cases have increased in 2018-2019. We can see that there were only a few cases from 2010-2017 which then rapidly increased in 2018 and 2019.

Calculating syphilis rates to compare with Oregon rates:
```{r}
st_year_wide<-spread(st_cases,month,count)
for (i in st_year_wide$year){
  st_year_wide$tot_cases <- rowSums(st_year_wide[2:13],na.rm=TRUE)
  print(st_year_wide)
}

## including another variable for total population
st_year_wide$tot.pop<- c(85735,85995,86785,87725,88740,90005,92575,93590,94360)
```
```{r}
#Note that the data for 2016 is unavailable. Adding 2016 as zero.
add_d<-data.frame("year"=as.double(2016),"1"=0,"2"=0,"3"=0,"4"=0,"5"=0,"6"=0,"7"=0,"8"=0,"9"=0,"10"=0,"11"=0,"12"=0,"tot_cases"=as.double(0),"tot.pop"=as.double(91320),check.names = FALSE)
add_d[2:13]<-as.integer(add_d[2:13])

## adding the 2016 data into the original dataset:
st_year<-merge(st_year_wide,add_d,all = TRUE)

## caluclating rates per 100,000:
st_year<- st_year%>%
  mutate(rate=(tot_cases/tot.pop)*100000)

## to view the dataset:
View(st_year)
```
Lets pull data for Oregon to compare rates with the Benton county. Same procedure is used to edit Oregon data:
```{r}
st_state<-read.csv("st-state.csv",header=TRUE,as.is=TRUE,sep=",",na.strings = " ")

st_state$CollectionDate<-mdy(st_state$EXP_Lab..CollectionDate)
st_state = st_state %>% 
  mutate(CollectionDate = ymd(CollectionDate)) %>% 
  mutate_at(vars(CollectionDate), funs(year, month, day)) %>%
  filter(year>=2010 & year<=2019)

st_state_cases<- st_state%>%
  group_by(year,month)%>%
  summarise(OR_count=n())
st_state_cases
st_state_wide<-spread(st_state_cases,month,OR_count)
st_state_wide
```
State data contains data for 2016. Do I have to remove it so that I can compare it with the Benton county data?- No , dont exclude any data..instead add 2016 to Benton but let the value be zero.
```{r}
for (i in st_state_wide$year){
  st_state_wide$OR_cases <- rowSums(st_state_wide[2:13],na.rm=TRUE)
  print(st_state_wide)
}

st_state_wide$tot_pop<- c(3837300,	3857625,	3883735,	3919020,	3962710,	4013845,4076350,	4141100,	4195300,	4236400)
st_state_wide<-st_state_wide%>%
  mutate(OR_rate=(OR_cases/tot_pop)*100000)
View(st_state_wide)
```
Joining both Oregon rates with Benton data:
```{r}
st_year$OR_cases<-st_comp$OR_cases
st_year$OR_rates<-st_comp$OR_rate

View(st_year)
```
5- year average (2015-2019):
```{r}
st_five_avg<- st_year %>%
  filter(year<=2019 & year>=2015)
mean(st_five_avg$tot_cases)
mean(st_five_avg$rate)
mean(st_five_avg$OR_cases)
mean(st_five_avg$OR_rates)
```
10- year average (2010-2019):
```{r}
mean(st_year$tot_cases)
mean(st_year$rate)
mean(st_year$OR_cases)
mean(st_year$OR_rates)
```
Plot for Benton cases per year:
```{r}
st_year$year<-as.factor(st_year$year)
ggplot(st_year,aes(x=year,y=tot_cases,group=1))+
  geom_line(aes(x=year,y=tot_cases,group=1,colour="Benton"),colour="blue",lwd=2)+
  geom_point(colour="blue",lwd=3)+
  scale_y_continuous()+
  theme_bw()+
  labs(x="Year", y="Number of cases", title = "Syphilis cases in Benton County between 2010 and 2019")
```
Cases per year in Oregon:
```{r}
ggplot(st_year,aes(x=year,y=OR_cases,group=1))+
  geom_line(colour="dark green",lwd=2)+
  geom_point(colour="dark green",lwd=3)+
  scale_y_continuous()+
  theme_bw()+
  labs(x="Year", y="Number of cases", title = "Syphilis in Oregon from 2010-2019")
```
Now lets plot Benton and OR rates:
```{r}
st_year$year<-as.factor(st_year$year)
colors<-c("Benton county"="blue","Oregon"="darkgreen")
ggplot(st_year,aes(x=year,y=rate,group=1))+
  geom_line(aes(x=year,y=rate,group=1,colour="Benton"),lwd=2)+
  geom_line(aes(x=year,y=OR_rates,group=1,colour="Oregon"),lwd=2)+
  geom_point(aes(x=year,y=rate,group=1),colour="blue",lwd=3)+
  geom_point(aes(x=year,y=OR_rates,group=1), colour="darkgreen",lwd=3)+
  scale_y_continuous()+
  theme_bw()+
  labs(x="Year", y="Rate (per 100,000)", title = "Syphilis rates in Benton County and Oregon- 2010 to 2019",color="Legend")+
  scale_colour_manual(name="Legend",values=c(Benton="blue", Oregon="darkgreen"))
```
The plot comparing rates show that Benton county has way more rates than Oregon in 2018 and 2019 although rates in both the places have increased in that period.

Gender data for Benton:

I am adding data for 2016 and 2014 for presentation and comparison purposes. 
```{r}
# creating new dataset including year, Sex, and count of cases
st_gender<- st%>%
  group_by(year,Sex)%>%
  summarise(number=n())

# adding 2016 and 2014 data as zero:
add_s<-data.frame("year"=as.double(2016),"Sex"=as.character(c("","F","M")),"number"=as.integer(c(0,0,0)),check.names = FALSE)
add_s1<-data.frame("year"=as.double(2014),"Sex"=as.character(c("F","M")),"number"=as.integer(c(0,0)),check.names = FALSE)

# merging the above created data with the original dataset created for gender:
st_gender_f<-merge(st_gender,add_s,all =TRUE)
st_gender_f<-merge(st_gender_f,add_s1,all=TRUE)
st_gender_f
# wide format:
st_gender_wide<-spread(st_gender_f,year,number)
st_gender_wide

# adding a new variable for total cases by gender
for (i in st_gender_wide$Sex){
  st_gender_wide$cases<- rowSums(st_gender_wide[2:11],na.rm=TRUE)
  print(st_gender_wide)
}

st_gender_f$year<-as.factor(st_gender_f$year) 
```

Pyramid plot:
```{r}
# st_gender_f$year<-as.factor(st_gender_f$year)
ggplot(data=st_gender_f) +
  geom_bar(aes(year,number,group=Sex,fill=Sex), stat = "identity",subset(st_gender,Sex=="F")) +
  geom_bar(aes(year,-number,group=Sex,fill=Sex), stat = "identity",subset(st_gender,Sex=="M")) +
  scale_y_continuous(breaks=seq(-50,50,2),labels=abs(seq(-50,50,2))) +
  scale_x_continuous(breaks=seq(2010,2019,1),labels=seq(2010,2019,1))+
  theme_bw()+
  labs(x="Year",y="Number of cases",title="Syphilis cases in Benton county by gender between 2010 and 2019")+
  coord_flip() +
  scale_fill_discrete(name="Gender", breaks=c("F", "M"),labels=c("Female","Male"))
```
*NO NEED TO PRESENT A PYRAMID PLOT FOR OREGON*
For Oregon,
```{r}
#st_state
st_state_gender<- st_state%>%
  group_by(year,Sex)%>%
  summarise(number=n())
st_state_gender<- st_state_gender[which(st_state_gender$year<2020 & st_state_gender$year>=2010),]
state_gender_wide<-spread(st_state_gender,year,number)
#state_gender_wide
for (i in state_gender_wide$Sex){
  state_gender_wide$cases <- rowSums(state_gender_wide[2:10],na.rm=TRUE)
  print(state_gender_wide)
}
#st_state_gender
```
Pyramid plot:
```{r}
ggplot(data=st_state_gender) +
  geom_bar(aes(year,number,group=Sex,fill=Sex), stat = "identity",subset(st_state_gender,Sex=="F")) +
  geom_bar(aes(year,-number,group=Sex,fill=Sex), stat = "identity",subset(st_state_gender,Sex=="M")) +
  scale_y_continuous(breaks=seq(-1000,1000,50),labels=abs(seq(-1000,1000,50))) +
  scale_x_continuous(breaks= seq(2010,2020,1))+
  theme_bw()+
  labs(x="Year",y="Number of cases",title="Cases per year for males and females")+
  coord_flip() +
  scale_fill_discrete(name="Gender", breaks=c("F", "M"),labels=c("Female","Male"))
```
lets work on a normal plot with two lines for males and females:
looks like we need to add 2016 data as zero for comparison purposes,
```{r}
st_gender_f[which(st_gender_f$Sex=="" ),]<-NA
st_gender_f<- na.omit(st_gender_f)
st_gender_f$year<-as.factor(st_gender_f$year)
ggplot(st_gender_f,aes(x=year,y=number,group=Sex))+
  geom_line(aes(col=Sex),lwd=2)+
  geom_point(aes(col=Sex),lwd=2)+
  labs(x="Year",y="Number of cases",title="Syphilis cases in Benton county by gender between 2010-2019")+
  scale_colour_discrete(name="Gender", breaks=c("F", "M"),labels=c("Female","Male"))+
  theme_bw()
```
Now lets try to graph the same for the state:
```{r}
#table(st_state_gender$Sex)
st_state_gender[which(st_state_gender$Sex=="" ),]<-NA
st_state_gender<- na.omit(st_state_gender)
st_state_gender$year<-as.factor(st_state_gender$year)
ggplot(st_state_gender,aes(x=year,y=number,group=Sex))+
  geom_line(aes(col=Sex),lwd=2)+
  geom_point(aes(col=Sex),lwd=2)+
  labs(x="Year",y="Number of cases",title="Number of cases in Oregon between 2010-2019 categorized by gender")+
  scale_colour_discrete(name="Gender", breaks=c("F", "M","X"),labels=c("Female","Male","X"))+
  theme_bw()
```
Cases according to age group:
```{r}
# creating new dataset for age, year and sex:
st_age_cat<- st%>%
  group_by(Age,Sex)%>%
  summarise(counts=n())

# Categorising age into groups:
st_age_cat<- st_age_cat%>%
  mutate(age_cat=case_when(Age < 10 ~ "(0-9)",
                           Age >= 10  & Age <= 19 ~ "(10-19)",
                           Age >= 20  & Age <= 29 ~ "(20-29)",
                           Age >= 30  & Age <= 39 ~ "(30-39)",
                           Age >= 40  & Age <= 49 ~ "(40-49)",
                           Age >= 50  & Age <= 59 ~ "(50-59)",
                           Age >= 60  & Age <= 69 ~ "(60-69)",
                           Age >= 70  & Age <= 79 ~ "(70-79)",
                           Age >= 80  & Age <= 89 ~ "(80-89)",
                           Age >= 90  & Age <= 99 ~ "(90-99)"))

st_age_cat # view the data
```
Graphs:
```{r}
# removing missing values:
st_age_cat<-na.omit(st_age_cat)
st_age_cat$counts<-as.integer(st_age_cat$counts,na.rm=TRUE)
ggplot(st_age_cat,aes(counts,group=Sex,fill=Sex))+
  geom_bar(aes(age_cat,counts,group=Sex,fill=Sex), stat = "identity",subset(ct_age_cat,Sex=="F"),na.rm=TRUE) +
  geom_bar(aes(age_cat,-counts,group=Sex,fill=Sex), stat = "identity",subset(ct_age_cat,Sex=="M"),na.rm=TRUE) +
  scale_y_continuous(breaks=seq(-1000,4000,200),labels=abs(seq(-1000,4000,200))) +
  theme_bw()+
  labs(x="Age range",y="Number of cases",title="Syphilis in Benton county by age group and gender (2010-2019)")+
  coord_flip()+
  scale_fill_discrete(name="Gender", breaks=c("F", "M"),labels=c("Female","Male"))
```
Cases according to Race:
```{r}
#creating new dataset for race, year and number of cases per each race for every year:
race_st_cases<- st%>%
  group_by(year,EXP_Patient..Race)%>%
  summarise(count=n())
```

```{r}
table(race_st_cases$EXP_Patient..Race)

#creating new category combining categories with two or more races:
race_st_cases$EXP_Patient..Race[which(race_st_cases$EXP_Patient..Race=="Asian\nOther"|race_st_cases$EXP_Patient..Race=="Other\nBlack\nWhite"| race_st_cases$EXP_Patient..Race=="White\nAI/AN")] <- "Two or more races"

race_st_cases$EXP_Patient..Race<-as.character(race_st_cases$EXP_Patient..Race)

# assigning and excluding NA's:
race_st_cases[which(race_st_cases$EXP_Patient..Race==""),]<- NA
race_st_cases<-na.omit(race_st_cases)

# adding new variable for total population:
race_st_cases<- race_st_cases%>%
  mutate(total_pop = case_when(year == "2010" ~ 85735 ,
                           year == "2011" ~ 85995,
                          year == "2012" ~86785,
                            year == "2013" ~87725,
                          year == "2014" ~88740,
                          year == "2015" ~90005,
                          year == "2016" ~91320,
                          year == "2017" ~92575,
                          year == "2018" ~93590,
                          year == "2019" ~94360))

# "rate" variable:
race_st_cases$rate<-c(((race_st_cases$count)/(race_st_cases$total_pop))*100000)

#view the edited data
race_st_cases
```
Graph:
```{r}
race_st_cases$year<-as.factor(race_st_cases$year)
ggplot(race_st_cases,aes(year,count,group=EXP_Patient..Race))+
  geom_line(aes(colour=EXP_Patient..Race),linetype=3)+
  geom_point(aes(colour=EXP_Patient..Race),lwd=2)+
  labs(x="Year",y="Number of cases",title="Syphilis cases in Benton county categorised by race (2010-2019)")+
  theme_bw()
```
*Footnote*
Unknown category in the Race variables include those people who did not know their race 

If I plot the graph without including those who did not provide their information?

For syphilis rates and year categorized by race,
```{r}
race_st_cases$year<-as.factor(race_st_cases$year)
ggplot(race_st_cases,aes(year,rate,group=EXP_Patient..Race))+
  geom_line(aes(colour=EXP_Patient..Race),linetype=3)+
  geom_point(aes(colour=EXP_Patient..Race),lwd=2)+
  labs(x="Year",y="rates",title="Syphilis rates (per 100,000) in Benton county by Race- 2010-2019")+
  theme_bw()
```
Seasonal variation of cases in Benton county:

Spring (March-may)
Summer (June-august)
Fall (September-november)
Winter (December- february)

Lets group data according the seasons:
```{r}
# creating tnew dataset with year, season, and sum of number of cases
st_season<-st_cases%>%
  mutate(season_cases= ifelse((month==12|month==1|month==2),"Winter",ifelse((month==3|month==4|month==5),"Spring",ifelse((month==6|month==7|month==8),"Summer",ifelse((month==9|month==10|month==11),"Fall","NA")))))%>%
  group_by(year,season_cases)%>%
  summarise(st_tot= sum(count))
st_season
```
Now that we have seasonal level count, we will plot the cases grouped by seasons:
```{r}
# assigning missing values and excluding them:
st_season[which(st_season$season_cases=="NA"),]<-NA
st_season<-na.omit(st_season)
```
Barplot for cases by year and season:
```{r}
st_season$year<-as.factor(st_season$year)
ggplot(data=st_season,mapping=aes(x=year,y=st_tot,group=season_cases,fill=season_cases))+
  geom_bar(stat="sum",na.rm=T,position = "stack",colour="white",lwd=0.5)+
  geom_text(aes(label=st_tot),position = "stack",size=2.5,colour="white")+
  #scale_y_continuous(breaks=c(0,600,100), labels=c(0,600,100))+
  theme_bw()+
  labs(x="Year",y="Number of cases ",title="Syphilis cases in Benton county by season (2010-2019)")+
  scale_fill_discrete(name="Seasons")
```
